Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2022-02-21 write-up/letters/response-ref-1-v1.tex

\vspace{5mm}
Thank you for raising this issue.  We discuss reporting delays and the relevant literature in Remark 1 in Section 2.2.  Given currently available public data, one cannot correct at the daily level for reporting delays.  Instead, we analyze COVID-19 data at the weekly level to account for potential reporting delays.  While not a perfect solution, it avoids clear reporting variation across day-of-week which can be seen in Figure 5.
\vspace{5mm}
\end{enumerate}
\newpage
{\bf Response to Reviewer 1}
\vspace{5mm}

I am incredibly grateful for this review. (PRAISE OUT THE WAZZOOO).

\begin{enumerate}
\item {\bf Section 1: Introduction.} {\it It may be useful in the introduction to briefly discuss the various purposes of testing for diseases – for COVID we have largely focused on diagnostic testing of people who are symptomatic or have a known or suspect exposure. Diagnostic testing will never provide valid prevalence or incidence estimates, especially for a disease where a significant proportion of cases are asymptomatic or pauci-symptomatic and thus unlikely to present for diagnostic testing. Screening tests are more appropriate as a potential tool for reconstructing prevalence or incidence tests, although these are seldom conducted at random. The random and non-random testing samples appear to represent screening and diagnostic testing goals, respectively, and this is part of why they can be expected not to deliver the same prevalence estimates.}

\vspace{5mm}
In the introduction, we now connect random testing to its purpose as a screening tool and non-random testing which generates the case count data as a diagnostic tool.  Given the infrequent nature of random testing and desire for ongoing prevalence estimates, we motivate the second objective of our paper by the question of whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing to estimate prevalence or incidence over time.
\vspace{5mm}

@@

	%https://academic.oup.com/aje/article/189/11/1438/5864525?searchresult=1, plot SIR
	%https://academic.oup.com/aje/article/178/9/1505/89262?searchresult=1, CORI
	%https://www.nature.com/articles/s41598-020-76563-8.pdf?origin=ppub, Althaus approximate

	This estimator is originally motivated by the SIR model and is exact under the poisson likelihood; however, the moment-based estimator has been proposed as a general tool for assessing epidemic trends in real time (Xinhua Yu, Assessing Epidemic Trends in Real Time With a Simple Ratio Plot, Am J of Epid, 189(11), 2020).  The estimator is closely related to the instantaneous reproduction numbers proposed by Cori et. al. (2013), but we now rely on recent work (Heng and Althaus, The approximately universal shapes of epidemic curves in the SEIR model, Sci Reports, 2020) to propose an approximate moment-based estimator in the SEIR setting. Specifically, Heng and Althaus (2020) derive the following approximate formula for the basic reproduction number under SEIR dynamics:
	$$
	R_0 = \frac{1 + \Lambda \left(\frac{1}{\sigma} + \frac{1}{\gamma} \right)}{S_0 + I_0 (1 + \gamma/\sigma)}.
	$$
	where~$\Lambda$ is the epidemic growth rate, and $S_0$/$I_0$ are the initial fractions of the population that are susceptible and infectious respectively.  For $S_0 \approx 1$ and $I_0 \ll 1$, we have the approximate relationship $R_0 \approx 1 + (\frac{1}{\gamma} + \frac{1}{\sigma})\Lambda$.
	, this is an approximate solution which lacks the term $\Lambda^2 \frac{1}{\gamma \sigma}$.  ``It corresponds to the case of an exponentially distributed generation time with mean duration $1/\gamma + 1/\sigma$, which is the same as the solution for the SIR model assuming an infectious period of $1/\gamma + 1/\sigma$. For realistic growth rates and choices of $\sigma$ and $\gamma$ for the COVID-19 pandemic, this term is negligible and therefore we work with the approximate formula.
	\vspace{5mm}

	\noindent Note that for the SIR model, the basic reproduction number is~$R_0 = 1 + \frac{1}{\gamma} \Lambda$. So estimate of the instantaneous reproduction number is obatined by replacing $\Lambda$ by an estimate of the  instantaneous growth rate $\hat \Lambda_t = \log (K_t / K_{t-1})$.  Then using the approximate relationship between the SEIR model and an SIR model with infectious period $1/\gamma + 1/\sigma$, we arrive at the new moment-based estimator
	$$
    R_t \approx 1 + \left( \frac{1}{\gamma} + \frac{1}{\sigma} \right) \log \left( \frac{K_t}{K_{t-1}} \right).
	$$
	The updated $R_t$ estimator shows that the same intuition about the error holds when we are considering an SEIR rather than an SIR model.  Figure 3b shows the reproduction number estimates and potential biases under SEIR dynamics which shows the same general conclusions hold.  For completeness, we also present the instantaneous reproduction number estimates using the proposal of Cori et. al. (2014) in Section I of the supplementary materials, which shows the same general conclusions.
\end{itemize}
\item {\bf Section 3.4: Rate comparisons}
\begin{itemize}
	\item {\it This section would also be more relevant with an SEIR model}
	\vspace{5mm}

### 2022-02-21 write-up/covid-writeup.tex
Intro updated

Due to monetary and time constraints, however, random testing is performed infrequently.  As of June 2021, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{These are the only random samples to collect both seroprevalence and diagnostic testing results. The CDC and other states have conducted seroprevalence-only studies.}. Indiana's sample was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

@@

Here we provide the necessary background to understand COVID-19 diagnostic and serological testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.

\subsection{Purposes of testing}

>>>>

Selection bias in case-count data is primarily due to it being a \emph{diagnostic tool}, i.e., individuals who are symptomatic or have a known/suspected exposure are more likely to present for diagnostic testing.  Case-count data arising from non-random testing means it cannot provide valid prevalence or incidence estimates due to the significant proportion of asymptomatic and pauci-symptomatic cases. Random testing, on the other hand, is used for \emph{screening purposes}, i.e., is an appropriate tool for reconstructing prevalence/incidence estimates. Due to monetary and time constraints, however, random testing is performed infrequently.  As of June 2021, Indiana and Ohio are the only states to conduct statewide random sample testing\footnote{These are the only random samples to collect both seroprevalence and diagnostic testing results. The CDC and other states have conducted seroprevalence-only studies.}. Indiana's sample was collected from April 25--29, 2020~\citep{Yiannoutsos2021}.  Such infrequent random testing is likely to provide insufficient information to help researchers and policy makers better understand the disease trajectory which can change rapidly over time.  Therefore, while random testing may be preferable in theory, in practice governments, researchers, and policy makers continue to use coronavirus case-counts to understand the impact of COVID-19 on the population and make data-informed decisions.

% Given the infrequent nature of random testing, however, a question is whether randomized screening tests can be used to address selection bias in the non-random diagnostic testing.

@@

Here we provide the necessary background to understand COVID-19 diagnostic testing, its scientific use in managing the pandemic, and the data streams considered in this manuscript.
